Quantum H theorem, variational entropy, and relaxation processes

### 2020-12-23 tex/classical.tex
Reworded: classical section finished (1st round)

Classical statistical mechanics is theoretically founded on the Boltzmann transport
equation and the hypo of molecular chaos. The fundamental connection
between the microscopic nature of matter and the phenomenological laws of thermodynamics
is the $H$-theorem.

@@

@@

As we saw in the previous section,
the validity of the $H$-theorem relies significantly on assuming
that the distribution function is homogeneous and that the hypothesis of
molecular chaos is surmised. However, in order to consider systems whose
distribution functions are not necessarily homogeneous, which
allows for studying systems out of equilibrium,
we introduce a modified $H$-functional as follows. Hereafter, we will
use the term homogeneous distribution function to refer to a distribution
function thad does not depend on the coordinates, and will use primed functions
and quantities to differentiate the classical case (this Section) \textit{vs} the quantum
case (below, Section \ref{sec:quantum}).

@@

If the complete system is in equilibrium without external forces applied to
the gas, from classical thermodynamics of
systems in equilibrium, we ascertain that the local number
of particles and the local
energy do not depend on the cell number, this is
%
\begin{subequations}\label{eq:eqNandE}
\begin{equation}
   \mathcal{N}'_M=\mathcal{N}'\equiv \bar{\mathcal{N}}'
\end{equation}
%
and
%
\begin{equation}
   \quad\mathcal{E}'_M=\mathcal{E}'\equiv\bar{\mathcal{E}}'.
\end{equation}
\end{subequations}
%
In Eqs. (\ref{eq:eqNandE}) the bar implies averaged properties over the complete system.
Moreover, the distribution function is homogeneous, hence $f'_M$
does not depend on the cell number $M$ (\textit{i.e.} $f'_M(\vec v,t)=f'(\vec v,t)$,
$\forall M$). Several properties follow directly from this, \textit{e.g.}
from Eqs.~(\ref{eq:eqNandE}) $E=\sum_M\mathcal{E}'_M=K\bar{\mathcal{E}}'$, $N=\sum_M\mathcal{N}'_M=K\bar{\mathcal{N}}'$ (here we have used Eqs. (\ref{eq:Ecell})
and (\ref{eq:Ncell})),
and Eq.~(\ref{eq:CH2}), in terms of Eq.~(\ref{eq:hbfunctional}), reduces to
%
\begin{equation}
    \mathcal{H}'(t)=\int \sum_{M=1}^{K} [f'(\vec{v},t)\ln f'(\vec{v},t)] \mathrm{d}^{3}v
         = K\int  f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v= K H_{B}(t).
\end{equation}
%
In order to identify $\mathcal{H}'$ with the entropy, it remains to show that $\mathcal{H}'(t)$ is extensive,
with respect to $Kf'_(\vec v,t)$.
This is shown by analyzing the following expression
%
\begin{eqnarray}\label{eq:sum-h}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v & = & 
        \int \big[(K\ln K)f'(\vec{v},t) + Kf'(\vec{v},t) \ln f'(\vec{v},t)\big]\mathrm{d}^{3}v\nonumber\\
        & = & K\int\Big\{ f'(\vec{v},t)\big[\ln K + \ln f'(\vec{v},t)\big] \Big\}\mathrm{d}^{3}v.
\end{eqnarray}
%
We observe that if the number of particles in the $\mu$-space, $f'(\vec{v},t)$, is much greater than
the number of cells $K$, then the first term of Eq.~(\ref{eq:sum-h}) is negligible, and consequently:
%
\begin{equation}\label{aditive-property}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v \simeq
    K\int f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v,
\end{equation}
%
\textit{i.e.} $\mathcal{H}'_M$ is extensive, and the sum $\sum_M\mathcal{H}'_M$ is the $H$-functional of the complete
system, which reduces to the Boltzmann $H$-functional. Hence, $\mathcal{H}'$ can also be identified with the
system's entropy.

@@

We commence by differentiating Eq.~(\ref{eq:CH2}) with respect to time,
%
\begin{equation}\label{eq:dH1}
    \frac{d\mathcal{H}'}{dt}=\sum_{M=1}^{K}\int\left[
      1+\ln f'_M(\vec{v},t)
    \right]\dot f'_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
(Here and throughout the article, we use the standard notation $\dot h\equiv(dh/dt)$.)
Substituting Eq.~(\ref{eq:firstorder}) into Eq.~(\ref{eq:dH1}) yields
%
\begin{equation}\label{eq:dH1-1}
    \frac{d\mathcal{H}'}{dt}=\sum_{M=1}^{K}\int\bar f'(\vec{v}) \left[
      1+\ln \left\{
        \bar f'(\vec{v})+\bar f'(\vec{v})g'_M(\vec{v},t)
      \right\}
    \right]\dot g'_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
The logarithmic term of Eq.~(\ref{eq:dH1-1}) expanded up to the first-order term
of its Taylor series, around $g'_{M}(\vec{v},t)=0$, is
%
\begin{equation}\label{lnapproximationclassical}
    \ln [\bar{f}'(\vec{v})+\bar{f}'(\vec{v}) g'_{M}(\vec{v},t)] \approx
    \ln [\bar{f}'(\vec{v})]+ g'_{M}(\vec{v},t),
\end{equation}
%
and substituting this into Eq.~(\ref{eq:dH1-1}) gives
%
\begin{eqnarray}
    \frac{d\mathcal{H}'}{dt}&=&\sum_{M=1}^{K} \int \bar f'(\vec{v})\left[
      1+\ln \bar f'(\vec{v})+g'_M(\vec{v},t)
    \right]\dot g'_M(\vec{v},t)\mathrm{d}^3v.
\end{eqnarray}
%

So far, we have not restricted the initial condition of the system to be
of any particular form, other than being close to equilibrium. Assuming
that the initial state of the system is also such that the Lagrange multipliers
are homogeneous (\textit{i.e.} $\alpha_M=\alpha$ and $\beta_M=\beta$, for all $M$),
we have from Eq.~(\ref{eq:relacion1}):
\commr{[Prof. Carrillo: A esto no le encuentro razón física, salvo que es lo que
hace Huang\dots]}

@@

In the first term of Eq.~(\ref{eq:dH1-2}), we have used Eq.~(\ref{eq:fbardef}).

@@

Therefore the first term of Eq.~(\ref{eq:dH1-2}) is null because of
Eqs.~(\ref{eq:classsumdotdeltaseq0}) and (\ref{eq:sumderdeltaseq0}), which renders

@@

We sort the previous ideas in the following form
%
\begin{equation}\label{separacionclassical}
\begin{array}{rl}
  g^{+}_{J}=+|g^{+}_{J}|; & \dot{g}^{+}_{J}=-|\dot{g}^{+}_{J}|,\\
  g^{-}_{J}=-|g^{-}_{J}|; & \dot{g}^{-}_{J}=+|\dot{g}^{-}_{J}|,
 \end{array}
\end{equation}
%
and consequently, (\ref{classicalH3}) obtains the following form
\end{comment}
%-----------------------------
%
If the system's initial state is sufficiently close to equilibrium, it is
physically suitable
to assume that $\left|g'_M(\vec{v},t_0)\right|\to 0$ as $t\to\infty$ in a monotonous manner,
thus ${{\dot{g}}'}_I\phantom{.}\!^{+}(\vec{v},t)\leq0$ and
${{\dot{g}}'}_J\phantom{.}\!^{-}(\vec{v},t)>0$, for $t\geq t_0$. Consequently,
Eq.~(\ref{eq:classicalH3}) can be re-written as
%
\begin{equation}\label{eq:classicalH4}
    \frac{d\mathcal{H}'}{dt}=-\left[
      \sum_J^{L}\int\bar f(\vec{v})|g_J^{+}(\vec{v},t)|
        |\dot g_J^{+}(\vec{v},t)|\mathrm{d}^3v
      +\sum_J^{P}\int\bar f(\vec{v})|g_J^{-}(\vec{v},t)|
      |\dot g_J^{-}(\vec{v},t)|\mathrm{d}^3v 
    \right].
\end{equation}
%
Since every integrand in Eq.~(\ref{eq:classicalH4}) is positive,
for all $t$ and $\vec v$, and
$\left|g'_M(\vec{v},t_0)\right|\to 0$ as $t\to\infty$, it follows that
%
\begin{equation}\label{eq:dHpdtleq0}
 \frac{d\mathcal{H}'}{dt}\leq0.
\end{equation}
%

We can observe that $\bar{f}(\vec{v})$ in (\ref{eq:classicalH4}) is always
positive. Also the deviation and its derivative are positive, then all the
expression is positive. However, the global sign makes the derivative of the
$\mathcal{H}'$ with respect to time is always less to zero. With this we proved that
$\frac{d\mathcal{H}'}{dt}<0$. If the system is in equilibrium,
$g_{J}(\vec{v},t)=\dot g_J(\vec{v},t)=0$, therefore $\frac{d\mathcal{H}'}{dt}=0$.

\commr{CHECKPOINT}

Joining both results, we can say the following statement:
Consider a classical gas in a total volume $V$ (divided in $K$ cells of equal
volume elements), total energy $E$ and total number of free particles $N$.
Consider also the system has inhomogeneities and suffers a relaxation process.
If we define the following functional
%
\begin{equation}\label{CH3}
   \mathcal{H}'(t)=\sum_{M=1}^{K}\int f_M(\vec{v},t) \ln f_M(\vec{v},t)\mathrm{d}^3v,
\end{equation}
%
where $f_M(\vec{v},t)$ is the local distribution function of each cell in the
system, then in the first-order approximation
%
\begin{equation}
    \frac{d\mathcal{H}'}{dt} \leq 0.
\end{equation}
%
This statement will be the classical $H$-theorem with inhomogeneities.

In the next section, we present a proposal quantum version of the
$H$-theorem defined by Tolman and our proposal quantum version of the $H$
theorem applying the method of the volume divided into cells.

>>>>

Classical statistical mechanics is theoretically built upon the Boltzmann transport
equation and the hypothesis of molecular chaos. The fundamental connection
between the microscopic nature of matter and the phenomenological laws of thermodynamics
is the $H$-theorem. In this article we propose an alternative approach, which is 
developed by means of the variational method, applied to a functional that may include
systems out-of-equilibrium (or non-homogeneous). Our proposal is limited to systems
whose initial state is physically close to the equilibrium, in such a manner that the
local equilibrium hypothesis is satisfied.

We begin this section by quite briefly stating the Boltzmann transport equation, followed
by an also quite brief review of the $H$-theorem demonstration, as performed in
standard textbooks.

@@

Before introducing our proposal, we bring the reader's attention to the following. In defining
Eq.~(\ref{eq:hbfunctional}), it is assumed that the distribution function does not
depend on the coordinates, which implies that the initial state of the system may not in
equilibrium, but the difference between the equilibrium and non-equilibrium is such
that it does not cause inhomogeneities to the system.

@@

As we saw in the previous section,
the validity of the $H$-theorem relies significantly on assuming
that the distribution function is homogeneous and that the hypothesis of
molecular chaos is surmised. However, in order to consider systems whose
distribution functions are not necessarily homogeneous, which
allows for studying systems out of equilibrium,
we introduce a modified $H$-functional as follows. Hereafter, we will
use the term homogeneous distribution function to refer to a distribution
function thad does not depend on the coordinates. In addition,
in this section we will use primed functions
and quantities to differentiate the classical case (this Section) \textit{vs} the quantum
case (below, Section \ref{sec:quantum}).

@@

If the complete system is in equilibrium without external forces applied to
the gas, from classical thermodynamics of
systems in equilibrium, we ascertain that the local number
of particles and the local
energy do not depend on the cell number, on a statistical sense, this is
%
\begin{subequations}\label{eq:eqNandE}
\begin{equation}
   \mathcal{N}'_M=\mathcal{N}'\equiv \bar{\mathcal{N}}'
\end{equation}
%
and
%
\begin{equation}
   \quad\mathcal{E}'_M=\mathcal{E}'\equiv\bar{\mathcal{E}}'.
\end{equation}
\end{subequations}
%
In Eqs. (\ref{eq:eqNandE}) the bar implies averaged properties over the complete system.
Moreover, the distribution function is homogeneous, hence $f'_M$
does not depend on the cell number $M$ (\textit{i.e.} $f'_M(\vec v,t)=f'(\vec v,t)$,
$\forall M$). Several properties follow directly from this, \textit{e.g.}
from Eqs.~(\ref{eq:eqNandE}) $E=\sum_M\mathcal{E}'_M=K\bar{\mathcal{E}}'$, $N=\sum_M\mathcal{N}'_M=K\bar{\mathcal{N}}'$ (here we have used Eqs. (\ref{eq:Ecell})
and (\ref{eq:Ncell})),
and Eq.~(\ref{eq:CH2}), in terms of Eq.~(\ref{eq:hbfunctional}), reduces to
%
\begin{equation}
    \mathcal{H}'(t)=\int \sum_{M=1}^{K} [f'(\vec{v},t)\ln f'(\vec{v},t)] \mathrm{d}^{3}v
         = K\int  f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v= K H_{B}(t).
\end{equation}
%
In order to identify $\mathcal{H}'$ with the entropy, it remains to show that $\mathcal{H}'(t)$ is extensive,
with respect to $Kf'_(\vec v,t)$.
This is shown by analyzing the following expression
%
\begin{eqnarray}\label{eq:sum-h}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v & = & 
        \int \big[(K\ln K)f'(\vec{v},t) + Kf'(\vec{v},t) \ln f'(\vec{v},t)\big]\mathrm{d}^{3}v\nonumber\\
        & = & K\int\Big\{ f'(\vec{v},t)\big[\ln K + \ln f'(\vec{v},t)\big] \Big\}\mathrm{d}^{3}v.
\end{eqnarray}
%
We observe that if the number of particles in the $\mu$-space, $f'(\vec{v},t)$, is much greater than
the number of cells $K$, then the first term of Eq.~(\ref{eq:sum-h}) is negligible, and consequently:
%
\begin{equation}\label{aditive-property}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v \simeq
    K\int f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v,
\end{equation}
%
\textit{i.e.} $\mathcal{H}'_M$ is extensive, and the sum $\sum_M\mathcal{H}'_M$ is the $H$-functional of the complete
system, which reduces to the Boltzmann $H$-functional. Hence, $\mathcal{H}'$ can also be identified with the
system's entropy.

@@

We commence by differentiating Eq.~(\ref{eq:CH2}) with respect to time,
\begin{equation}\label{eq:dH1}
    \frac{d\mathcal{H}'}{dt}=\sum_{M=1}^{K}\int\left[
      1+\ln f'_M(\vec{v},t)
    \right]\dot f'_M(\vec{v},t)\mathrm{d}^3v.
(Here and throughout the article, we use the standard notation $\dot h\equiv(dh/dt)$.)
Substituting Eq.~(\ref{eq:firstorder}) into Eq.~(\ref{eq:dH1}) yields
\begin{equation}\label{eq:dH1-1}
    \frac{d\mathcal{H}'}{dt}=\sum_{M=1}^{K}\int\bar f'(\vec{v}) \left[
      1+\ln \left\{
        \bar f'(\vec{v})+\bar f'(\vec{v})g'_M(\vec{v},t)
      \right\}
    \right]\dot g'_M(\vec{v},t)\mathrm{d}^3v.
The logarithmic term of Eq.~(\ref{eq:dH1-1}) expanded up to the first-order term
of its Taylor series, around $g'_{M}(\vec{v},t)=0$, is
\begin{equation}\label{lnapproximationclassical}
    \ln [\bar{f}'(\vec{v})+\bar{f}'(\vec{v}) g'_{M}(\vec{v},t)] \approx
    \ln [\bar{f}'(\vec{v})]+ g'_{M}(\vec{v},t),
and substituting this into Eq.~(\ref{eq:dH1-1}) gives
    \frac{d\mathcal{H}'}{dt}&=&\sum_{M=1}^{K} \int \bar f'(\vec{v})\left[
      1+\ln \bar f'(\vec{v})+g'_M(\vec{v},t)
    \right]\dot g'_M(\vec{v},t)\mathrm{d}^3v.
Substituting $\ln \bar f'(\vec{v})=\exp\left(\alpha+\beta \epsilon(\vec{v})\right)$,
see Eq.~(\ref{eq:fbardef}) and text following it, and omitting $C$ yields

@@

@@

Therefore, because of
Eqs.~(\ref{eq:classsumdotdeltaseq0}) and (\ref{eq:sumderdeltaseq0}), Eq.~(\ref{eq:dH1-2}) simplifies to

@@

We sort the previous ideas in the following form
%
\begin{equation}\label{separacionclassical}
\begin{array}{rl}
  g^{+}_{J}=+|g^{+}_{J}|; & \dot{g}^{+}_{J}=-|\dot{g}^{+}_{J}|,\\
  g^{-}_{J}=-|g^{-}_{J}|; & \dot{g}^{-}_{J}=+|\dot{g}^{-}_{J}|,
 \end{array}
\end{equation}
%
and consequently, (\ref{classicalH3}) obtains the following form

We can observe that $\bar{f}(\vec{v})$ in (\ref{eq:classicalH4}) is always
positive. Also the deviation and its derivative are positive, then all the
expression is positive. However, the global sign makes the derivative of the
$\mathcal{H}'$ with respect to time is always less to zero. With this we proved that
$\frac{d\mathcal{H}'}{dt}<0$. If the system is in equilibrium,
$g_{J}(\vec{v},t)=\dot g_J(\vec{v},t)=0$, therefore $\frac{d\mathcal{H}'}{dt}=0$.
\end{comment}
%-----------------------------
%
If the system's initial state is sufficiently close to equilibrium, it is
physically suitable
to assume that $\left|g'_M(\vec{v},t_0)\right|\to 0$ as $t\to\infty$ in a monotonous manner,
thus ${{\dot{g}}'}_I\phantom{.}\!^{+}(\vec{v},t)\leq0$ and
${{\dot{g}}'}_J\phantom{.}\!^{-}(\vec{v},t)>0$, for $t\geq t_0$. Consequently,
Eq.~(\ref{eq:classicalH3}) can be re-written as
%
\begin{equation}\label{eq:classicalH4}
    \frac{d\mathcal{H}'}{dt}=-\left[
      \sum_J^{L}\int\bar f(\vec{v})|g_J^{+}(\vec{v},t)|
        |\dot g_J^{+}(\vec{v},t)|\mathrm{d}^3v
      +\sum_J^{P}\int\bar f(\vec{v})|g_J^{-}(\vec{v},t)|
      |\dot g_J^{-}(\vec{v},t)|\mathrm{d}^3v 
    \right].
\end{equation}
%
Since every integrand in Eq.~(\ref{eq:classicalH4}) is positive,
for all $t$ and $\vec v$, and
$\left|g'_M(\vec{v},t_0)\right|\to 0$ as $t\to\infty$, it follows that
%
\begin{equation}\label{eq:dHpdtleq0}
 \frac{d\mathcal{H}'}{dt}\leq0.
\end{equation}
%

In summary, if a gas occupying a volume V (which is divided into $K$ small cells), with a
total energy $E$ and $N$ classical free particles, whose initial state is not in equilibrium,
but sufficiently close to equilibrium ---such that the local equilibrium hypothesis apply---, then the funcional
%
\begin{equation}\label{eq:CH3}
   \mathcal{H}'(t)=\sum_{M=1}^{K}\int f'_M(\vec{v},t) \ln f'_M(\vec{v},t)\mathrm{d}^3v
\end{equation}
%
---where $f'_M(\vec{v},t)$ is the cell distribution function---,
satisfies $d\mathcal{H}'/dt\leq0$, and the equality relation is attained at $t\to\infty$.
In Eq.~(\ref{eq:CH3}), $f'_M(\vec{v},t)$ is the Maxwell-Boltzmann distribution
function, which in general is different for different cells (\textit{i.e.} the complete system
can be non-homogeneous), and each $f'_M(\vec v,t)$ is compatible
with the cell number of particles, $\mathcal{N}'_M$, energy, $\mathcal{E}'_M$, temperature,
$T_M$, and Legendre multipliers $\alpha_M$ and $\beta_M$.

\begin{comment}
Joining both results, we can say the following statement:
Consider a classical gas in a total volume $V$ (divided in $K$ cells of equal
volume elements), total energy $E$ and total number of free particles $N$.
Consider also the system has inhomogeneities and suffers a relaxation process.
If we define the following functional
%
\begin{equation}\label{CH3}
   \mathcal{H}'(t)=\sum_{M=1}^{K}\int f_M(\vec{v},t) \ln f_M(\vec{v},t)\mathrm{d}^3v,
\end{equation}
%
where $f_M(\vec{v},t)$ is the local distribution function of each cell in the
system, then in the first-order approximation
%
\begin{equation}
    \frac{d\mathcal{H}'}{dt} \leq 0.
\end{equation}
%
This statement will be the classical $H$-theorem with inhomogeneities.

In the next section, we present a proposal quantum version of the
$H$-theorem defined by Tolman and our proposal quantum version of the $H$
theorem applying the method of the volume divided into cells.
\end{comment}
